Convective heat transport in stratified atmospheres at low and high Mach number

### 2017-02-23 tex/nusselt_paper.tex
fig 3 (re/nu) is now ~finalized (other than calculating the last few data points).  Intro reworked but rough.

%%%%% Body of the paper
\section{Introduction}
\refstepcounter{section}
\label{sec:intro}
Convective heat transport is essential
in the cores of high mass stars,
the envelopes of low mass stars, and the atmospheres of terrestrial and 
jovian planets.  These astrophysical systems have density stratifications
ranging from a few density scale heights (e.g. massive star cores)
up to 14 density scale heights in the convective envelopes of low
mass stars such as the Sun.
Atmospheric stratification presumably modifies the fundamental
dynamics and heat transport mechanisms of these atmospheres.
Understanding the fundamental
properties of compressible convection in stratified media is essential
to characterizing systems in astrophysics and planetary sciences.  
Numerical constraints have often restricted studies of compressible
convection to moderately high Mach number (Ma), appropriate to regions near 
the Sun's surface.  Few fundamental properties of
low Ma convection, which occurs in the deep solar interior, are known.

Early numerical experiments on stratified convection
in two \cite{graham1975, chan&all1982,
hurlburt&all1984, cattaneo&all1990} and three 
\cite{cattaneo&all1991, brummell&all1996} dimensions
revealed a number of basic properties in the moderate-to-high 
Ma regime. In the widely-studied \RB (hereafter RB) problem, 
upflows and downflows are symmetrical and
the conductive flux approaches zero in the convective interior.
These two hallmark characteristics of RB convection change
significantly when stratification is included.  Stratified convection 
exhibits narrow downflow lanes and broad upflow regions.
Furthermore, the \emph{entropy} gradient is negated by convection 
rather than the temperature gradient, and
a significant conductive flux can exist in the presence of
efficient convection.

In RB convection, there exist two primary control parameters: 
the Rayleigh number (Ra, the ratio of
buoyant driving to diffusive damping) and the Prandtl number 
(Pr, the ratio of viscous to thermal
diffusivity).  These numbers coupled with the aspect ratio of 
the physical domain and the boundary conditions
determine the dynamics of the convection.  In stratified atmospheres, 
in addition to specifying the equation of state and
fundamental properties of the gas, the two control parameters of 
RB convection are joined by the degree of
stratification across the domain and the characteristic 
Ma of the convective flows.  
Polytropically stratified atmospheres, such as those used in 
early studies \cite{graham1975, chan&all1982, hurlburt&all1984, 
cattaneo&all1990, cattaneo&all1991, brummell&all1996}, are an ideal extension of
RB convection into the stratified realm because the two additional 
control parameters are directly linked to
basic properties of the atmosphere.  The density stratification is 
set by the number of density scale heights (\nrho)
the atmosphere spans, and Ma is controlled 
by the superadiabatic excess ($\epsilon$),
the deviation of the polytropic index from the adiabatic polytropic 
index \cite{graham1975}.

In this letter we study the behavior of convective heat transport, 
quantified by the Nusselt number (Nu), in plane-parallel, 
two-dimensional, polytropically stratified atmospheres.  
We vary $\epsilon$ and Ra while holding $\nrho$, Pr, and the aspect ratio
constant.  We describe experimental methods in section 
\ref{sec:experiment}, including the construction of atmospheres, equations, and numerical methods.  
Results are described in section \ref{sec:results} and their implications are discussed
in section \ref{sec:discussion}.

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
We exmine the simplest stratified extension of RB by studying a
fluid composed of monatomic ideal gas particles with an adiabatic 
index of $\gamma = 5/3$ and whose equation of state is $P = R\rho T$. 
This is consistent with the approach used in earlier work.
The initial atmosphere is a plane-parallel polytrope in which 
the gravitational acceleration and conductive flux, 
$\bm{F}_{\text{cond,0}} = -\kappa \partial_z T_0$, do not vary with depth. To
achieve the latter condition, both $\kappa$ and $\partial_z T_0$ are constant.
Under these assumptions, satisfying hydrostatic 
equilibrium produces a stratification of
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(z_0 - z)^m, \\
T_0(z)    &= T_{t}(z_0 - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m = m_{\text{ad}} - \epsilon$ is the polytropic index.
The adiabatic polytropic index is $m_{\text{ad}} \equiv (\gamma-1)^{-1}$, and
the superadiabatic excess is $\epsilon$ which sets the scale 
of the entropy gradient ($\partial_z S_0 \propto -\epsilon$).
A significant advance of this work is the ability to study large 
and small values of $\epsilon$, as will be discussed.
Thermodynamic variables are nondimensionalized at the 
top of the atmosphere as  $P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, 
requiring $z_0 \equiv L_z + 1$ and $R = T_{t} = \rho_{t} = 1$.
By this choice, the non-dimensional length scale is the inverse 
temperature gradient scale and the timescale is the isothermal 
sound crossing time of this unit length.
The height $z$ increases upwards within $[0, L_{z}]$, 
where $L_{z} = e^{n_{\rho}/m} - 1$ is
determined by $n_\rho$ and $\epsilon$.
The characteristic timescale of convective dynamics
is related to the atmospheric buoyancy time, 
$t_{\text{b}} = \sqrt{L_z/g\epsilon}$, with $g = (m+1)$.
Throughout this letter, we use buoyancy time units and 
choose $n_{\rho} = 3$ such that the 
initial density varies by a factor of 20.
All atmospheres studied here have an aspect ratio of 4, 
such that $L_x = 4L_z$.

>>>>

%%%%% Body of the paper
\section{Introduction}
\refstepcounter{section}
\label{sec:intro}
Convective dynamics transport energy in stellar and planetary atmospheres.
In these objects, flows are compressible and
feel the atmospheric stratification.  While in some systems this stratification is
negligible, it is significant in regions such as
the convective envelope of the Sun, which spans 14 density scale heights.
In the bulk of these systems, flows are at very low Mach Number (Ma), but numerical
constrainst have restricted most studies of compressible convection to high Ma.
These studies provide important insight to low temperature, high Ma regions
near the Sun's surface, but few fundamental
properties of low Ma convection which characterize deeper motions
are known.

%Convective heat transport is essential
%in the cores of high mass stars,
%the envelopes of low mass stars, and the atmospheres of terrestrial and 
%jovian planets.  These astrophysical systems have density stratifications
%ranging from a few density scale heights (e.g. massive star cores)
%up to 14 density scale heights in the convective envelopes of low
%mass stars such as the Sun.
%Atmospheric stratification presumably modifies the fundamental
%dynamics and heat transport mechanisms of these atmospheres.
%Understanding the fundamental
%properties of compressible convection in stratified media is essential
%to characterizing systems in astrophysics and planetary sciences.  
%Numerical constraints have often restricted studies of compressible
%convection to moderately high Mach number (Ma), appropriate to regions near 
%the Sun's surface.  Few fundamental properties of
%low Ma convection, which occurs in the deep solar interior, are known.

In the widely-studied \RB (RB) problem of Boussinesq convection, 
upflows and downflows are symmetrical and
the conductive flux ($\propto \grad T$) approaches zero in the convective interior.
Early numerical experiments of moderate-to-high Ma compressible convection
in two \cite{graham1975, chan&all1982,
hurlburt&all1984, cattaneo&all1990} and three 
\cite{cattaneo&all1991, brummell&all1996} dimensions
revealed that these two hallmark characteristics of RB convection change
significantly when stratification is included.  Downflow lanes
become fast and narrow, and upflow lanes turn into broad, slow upwellings.
Furthermore, the \emph{entropy} gradient is negated by convection 
rather than the temperature gradient, and
a significant conductive flux can exist in the presence of
efficient convection.

In RB convection, there exist two primary dynamical control parameters: 
the Rayleigh number (Ra, the ratio of
buoyant driving to diffusive damping) and the Prandtl number 
(Pr, the ratio of viscous to thermal
diffusivity). These numbers control two useful
measures of turbulence in the evolved solution:
the Reynolds
number (Re, the strength of advection to viscous diffusion)
and the Peclet number (Pe = Re Pr).  Stratified atmospheres
with negative entropy gradients are unstable to convection,
and the magnitude of that entropy gradient, or the superadiabaticity,
joins Ra and Pr as an important control parameter.  The
Ma of the evolved solution is primarily controlled by 
the \emph{superadiabatic excess}, $\epsilon$, which 
sets the scale of the atmospheric entropy gradient \cite{graham1975}.

%In RB convection, there exist two primary control parameters: 
%the Rayleigh number (Ra, the ratio of
%buoyant driving to diffusive damping) and the Prandtl number 
%(Pr, the ratio of viscous to thermal
%diffusivity).  These numbers coupled with the aspect ratio of 
%the physical domain and the boundary conditions
%determine the dynamics of the convection.  In stratified atmospheres, 
%in addition to specifying the equation of state and
%fundamental properties of the gas, the two control parameters of 
%RB convection are joined by the degree of
%stratification across the domain and the characteristic 
%Ma of the convective flows.  
%Polytropically stratified atmospheres, such as those used in 
%early studies \cite{graham1975, chan&all1982, hurlburt&all1984, 
%cattaneo&all1990, cattaneo&all1991, brummell&all1996}, are an ideal extension of
%RB convection into the stratified realm because the two additional 
%control parameters are directly linked to
%basic properties of the atmosphere.  The density stratification is 
%set by the number of density scale heights (\nrho)
%the atmosphere spans, and Ma is controlled 
%by the superadiabatic excess ($\epsilon$),
%the deviation of the polytropic index from the adiabatic polytropic 
%index \cite{graham1975}.

In this letter we study the behavior of convective heat transport, 
quantified by the Nusselt number (Nu), in plane-parallel, 
two-dimensional, polytropically stratified atmospheres.  
We vary $\epsilon$ and Ra while holding Pr, aspect ratio, boundary conditions,
and initial atmospheric stratification
constant.  We describe experimental methods in section 
\ref{sec:experiment}, including the construction of atmospheres, equations, and numerical methods.  
Results are described in section \ref{sec:results} and their implications are discussed
in section \ref{sec:discussion}.

\section{Experiment} 
\refstepcounter{section}
\label{sec:experiment}
We examine the simplest stratified extension of RB by studying a
fluid composed of monatomic ideal gas particles with an adiabatic 
index of $\gamma = 5/3$ and whose equation of state is $P = R\rho T$. 
This is consistent with the approach used in earlier work.
The initial atmosphere is a plane-parallel polytrope in which 
the gravitational acceleration and conductive flux, 
$\bm{F}_{\text{cond,0}} = -\kappa \partial_z T_0$, do not vary with depth. To
achieve the latter condition, both $\kappa$ and $\partial_z T_0$ are constant.
Under these assumptions, satisfying hydrostatic 
equilibrium produces a stratification of
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{t}(z_0 - z)^m, \\
T_0(z)    &= T_{t}(z_0 - z),
\label{eqn:polytrope}
\end{split}
\end{equation}
where $m = m_{\text{ad}} - \epsilon$ is the polytropic index.
The adiabatic polytropic index is $m_{\text{ad}} \equiv (\gamma-1)^{-1}$, and
the superadiabatic excess is $\epsilon$ which sets the scale 
of the entropy gradient ($\partial_z S_0 \propto -\epsilon$).
A significant advance of this work is the ability to study large 
and small values of $\epsilon$, as will be discussed.
Thermodynamic variables are nondimensionalized at the 
top of the atmosphere as  $P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, 
requiring $z_0 \equiv L_z + 1$ and $R = T_{t} = \rho_{t} = 1$.
By this choice, the non-dimensional length scale is the inverse 
temperature gradient scale and the timescale is the isothermal 
sound crossing time of this unit length.
The height $z$ increases upwards within $[0, L_{z}]$, 
where $L_{z} = e^{n_{\rho}/m} - 1$ is
determined by $n_\rho$ and $\epsilon$.
The characteristic timescale of convective dynamics
is related to the atmospheric buoyancy time, 
$t_{\text{b}} = \sqrt{L_z/g\epsilon}$, with $g = (m+1)$.
Throughout this letter, we use buoyancy time units and 
choose $n_{\rho} = 3$ such that the 
initial density varies by a factor of 20.
All atmospheres studied here have an aspect ratio of 4, 
such that $L_x = 4L_z$.
